---
title: "L08 Coords"
subtitle: "Data Visualization (STAT 302)" 
author: "Olivia Harbison"

format:
  html:
    toc: true
    embed-resources: true
    code-fold: show
    link-external-newwindow: true
    fig-dpi: 300

execute:
  warning: false

from: markdown+emoji 
---

## Overview

The goal of this lab is to use an alternative coordinate system in `ggplot2` to build a plot. 

## Dataset

We'll be using the `mod_nba2014_15_advanced.txt` dataset --- add to your project's `/data` subdirectory. The `codebook_mod_nba2014_15_advanced.txt` provides a quick description of the variables in the dataset --- suggest adding it to the `/data` subdirectory as well. 

```{r}
#| label: load-pkgs-data

# load package(s)
library(tidyverse)

# load data
nba_dat <- read.delim("data/mod_nba2014_15_advanced.txt", sep = "|") |> janitor::clean_names()
```

## Exercise 

<!-- See the html instructions since they are complete with images and are the official instructions. -->

Using the `mod_nba2014_15.txt` dataset we will try to recreate/approximate the plot type featured in the [http://fivethirtyeight.com/](fivethirtyeight.com) article [Kawhi Leonard Is The Most Well-Rounded Elite Shooter Since Larry Bird](http://fivethirtyeight.com/features/kawhi-leonard-is-the-most-well-rounded-elite-shooter-since-larry-bird/) for any player of your choice for the 2014-2015 season. 

**Replicate the graphic below as closely as possible:**

### Start with data wrangling

When data wrangling we will need a helper function for creating the quartiles for players. Additionally, it will be useful to exclude players that played less than 10 games or played less than 5 minutes a game. That is, we only want to include "qualified" players.

```{r}
#| label: helper-function

# Returns quartile rank
quartile_rank <- function(x = 0:99) {
  # Set quartile
  quart_breaks <- c(
    -Inf,
    quantile(x,
      probs = c(.25, .5, .75),
      na.rm = TRUE
    ),
    Inf
  )
  cut(x = x, breaks = quart_breaks, labels = FALSE)
}
```


::: {.callout-tip icon=false}

## Solution

```{r}
#| label: data-wrangling

# Graphical data
nba_graph_dat <- nba_dat %>% 
  # filtering for qualified players
  filter(g >= 10, mp / g >= 5) %>% 
  # adding quantile ranks for each player
  mutate(
    ts_quant = quartile_rank(ts_perc),
    trb_quant = quartile_rank(trb_perc),
    dbpm_quant = quartile_rank(dbpm),
    ast_quant = quartile_rank(ast_perc),
    usg_quant = quartile_rank(usg_perc)
  ) %>% 
  # selecting only the relevant variables
  select(player, contains("_quant")) %>% 
  # making the data longer so it's tidy and we can use/graph it
  pivot_longer(
    cols = -player, 
    names_to = "variable", 
    values_to = "value"
    ) %>% 
  # arranging by alphabetical order of players
  arrange(player)

plot_df <- nba_graph_dat %>%
  filter(player == "Jamal Crawford") 
```

:::

### Build the plot

*Hints:*

- Image width is 8 in, height is 10 in --- use code chunk yaml
- center the image output --- use code chunk yaml
- shrink size down for display to 50% width --- use code chunk yaml
- All added text, except title, is done with `annotate()`
- set `axis.text` to blank in theme
- size 3 and 5 could be useful

::: {.callout-tip icon=false}

## Solution
```{r}
#| fig-width: 8
#| fig-height: 10
#| out-width: 50%
#| fig-align: center


ggplot(plot_df, aes(x = variable, y = value)) +
  geom_col(fill = "#F28291",
           width = 1) +
  geom_hline(aes(yintercept = y),
             data.frame(y = c(1:4)),
             linetype = "dotted") +
  scale_x_discrete(NULL,
                   limits = c(
                     "ts_quant",
                     "usg_quant",
                     "dbpm_quant",
                     "trb_quant",
                     "ast_quant"
                   )) +
  scale_y_continuous(NULL) +
  geom_segment(x = 0.5,
               y = 0,
               xend = 0.5,
               yend = 3) +
  geom_segment(x = 1.5,
               y = 0,
               xend = 1.5,
               yend = 4) +
  geom_segment(x = 2.5,
               y = 0,
               xend = 2.5,
               yend = 4) +
  geom_segment(x = 3.5,
               y = 0,
               xend = 3.5,
               yend = 1) +
  geom_segment(x = 4.5,
               y = 0,
               xend = 4.5,
               yend = 3) +
  theme(axis.text = element_blank()) +
  coord_polar() +
  annotate(
    geom = "text",
    x = rep(3, 4),
    y = 0.7:3.7,
    label = c("1st-25th", "26th-50th", "51st-75th", "76th-99th"),
    size = 3
  ) +
  labs(title = "Jamal Crawford",
       subtitle = "2014-15 Season") +
  annotate(
    geom = "text",
    label = c(
      "TRUE\nSHOOTING",
      "USAGE\nRATE",
      "DEFENSIVE\nBPM",
      "REBOUND\nRATE",
      "ASSIST\nRATE"
    ),
    x = 1:5,
    y = rep(5, 5),
    size = 5
  )

```
:::

## Challenge

<!-- See the html instructions since they are complete with images and are the official instructions. -->

**Challenge is optional for all students, but we recommend trying them out!** 

Continue the work of trying to recreate the graphic from the article. There is a pre built theme from `ggthemes` that might be useful. Your final plot should look as follows:
